Structure-Preserving Model-Reduction of Dissipative Hamiltonian Systems

### 2017-03-07 dissipative_hamiltonian/chapters/2.Hamil.tex
Numerical result added

Under general coordinate changes, equations of evolution do not necessarily take the form of (\ref{eq:2.1}). Let $(\mathbb R^{2n},\Omega)$ and $(\mathbb R^{2k},\Lambda)$ be two symplectic linear vector spaces. A linear transformation $\alpha :\mathbb R^{2n} \to \mathbb R^{2k}$ is called a \emph{symplectic transformation} \cite{Marsden:2010:IMS:1965128} if
\begin{equation}
	\Omega(x,y) = \Lambda(\alpha(x),\alpha(y)), \quad \text{for all } x,y\in \mathbb R^{2n}.
\end{equation}
Alternatively in matrix notation, $A\in \mathbb R^{2n\times 2k}$ is called a \emph{symplectic matrix} if
\begin{equation}
	A^T \mathbb{J}_{2n} A = \mathbb{J}_{2k},
\end{equation}
where the superscript $T$ represents transpose operator. As the symplectic 2-form is preserved under symplectic transformations, the form of the equations of evolution remains invariant \cite{Marsden:2010:IMS:1965128}. The \emph{symplectic inverse} of $A$, is a form of pseudo-inverse given by
\begin{equation}
	A^+ = \mathbb{\mathbb J}_{2k}^T A^T \mathbb J_{2n}.
\end{equation}
It is shown in \cite{Peng:2014di} that $A^+A = I_{2k}$ and that $(A^+)^T$ is a symplectic matrix. Further, one can easily check that $AA^+$ is idempotent and so is a projection operator onto the column span of $A$.

@@

\begin{proof}
Here we prove the theorem for the case where $\chi$ is a constant symmetric matrix, where condition (\ref{eq:2.8}) corresponds to $\chi$ being positive semi-definite. We refer the reader to \cite{Figotin:2006jy} for the proof of the general case. Consider the Hamiltonian system
\begin{subequations}
\begin{align}
		\label{eq:2.10.a} & \dot{z}(t) = \mathbb J_{2n} K^T f(t), \\
		\label{eq:2.10.b} & \partial_t \theta(t,x) = \partial_x^2 \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\chi}  f(t), \\
		\label{eq:2.10.c} & \partial_t \phi(t,x) = \theta(t,x),
\end{align}
\end{subequations}
together with the initial condition
\begin{equation} \label{eq:2.10.1}
	z(0) = z_0,\quad \theta(0,\cdot) = 0, \quad \phi(0,\cdot) = 0.
\end{equation}
Here $\theta$ and $\phi$ are vector valued functions in $\mathcal H^{2n}$, $\delta_0(s)$ is the Dirac's delta function, $\sqrt{ \chi}$ is the matrix square root of $\chi$ and $f$ is the solution to the equation
\begin{equation} \label{eq:2.11}
	f(t) + \sqrt{2} \cdot \sqrt{ \chi } \phi(t,0) = Kz(t).
\end{equation}
To show that the Hamiltonian system (\ref{eq:2.10.a}) to (\ref{eq:2.10.c}) is an extension to (\ref{eq:2.4}) in the sense discussed before, it is enough to show that the solution $f$ to the equation (\ref{eq:2.11}) also satisfies (\ref{eq:2.4.1}). Equations (\ref{eq:2.10.b}) and (\ref{eq:2.10.c}) is a forced wave, with the analytical solution
\begin{equation} \label{eq:2.12}
	\phi(t,x) = \frac {\sqrt 2} 2 \int_0^{t-|x|} \sqrt{\chi} f(s)\ ds,\quad \theta(t,x) = \frac{\sqrt 2}{2} \cdot \sqrt{\chi} f(t - |x|).
\end{equation}
\end{proof}
We can recover (\ref{eq:2.4.1}) by substituting (\ref{eq:2.12}) into (\ref{eq:2.11}). The extended Hamiltonian for the system (\ref{eq:2.10.a}) to (\ref{eq:2.10.c}) takes the quadratic from
\begin{equation}
	\tilde h(z,\phi,\theta) = \frac 1 2 \left( \| Kz - \phi(t,0) \|_2^2 + \| \theta(t) \|^2_{\mathcal H^{2n} } + \| \partial_x\phi(t)\|^2_{\mathcal H^{2n} }\right)
\end{equation}
where $\| \cdot \|_2$ is the Euclidean norm on $\mathbb R^{2n}$ and $\| \cdot \|_{\mathcal H^{2n}}$ is the induced norm from the inner product on $\mathcal H^{2n}$.

@@

>>>>

Under general coordinate changes, equations of evolution do not necessarily take the form of (\ref{eq:2.1}). Let $(\mathbb R^{2n},\Omega)$ and $(\mathbb R^{2k},\Lambda)$ be two symplectic linear vector spaces. A linear transformation $\alpha :\mathbb R^{2n} \to \mathbb R^{2k}$ is called a \emph{symplectic transformation} \cite{Marsden:2010:IMS:1965128} if
\begin{equation}
	\Omega(x,y) = \Lambda(\alpha(x),\alpha(y)), \quad \text{for all } x,y\in \mathbb R^{2n}.
\end{equation}
Alternatively in matrix notation, $A\in \mathbb R^{2n\times 2k}$ is called a \emph{symplectic matrix} if
\begin{equation}
	A^T \mathbb{J}_{2n} A = \mathbb{J}_{2k},
\end{equation}
where the superscript $T$ represents the transpose operator. As the symplectic 2-form is preserved under symplectic transformations, the form of the equations of evolution remains invariant \cite{Marsden:2010:IMS:1965128}. The \emph{symplectic inverse} of $A$, is a form of pseudo-inverse given by
\begin{equation}
	A^+ = \mathbb{\mathbb J}_{2k}^T A^T \mathbb J_{2n}.
\end{equation}
It is shown in \cite{Peng:2014di} that $A^+A = I_{2k}$ and that $(A^+)^T$ is a symplectic matrix. Further, one can easily check that $AA^+$ is idempotent and so is a projection operator onto the column span of $A$.

It is known that symplectic matrices are usually ill-conditioned \cite{Karow:2006cf}. Under some conditions on a symplectic space \cite{da2003introduction}, one can construct a symplectic basis which is also ortho-normal, and thus norm bounded. A basis which is both symplectic and orthogonal is referred to as an \emph{ortho-symplectic basis}. We refer the reader to \cite{da2003introduction} for conditions on existence and construction of an ortho-symplectic basis.

@@

\begin{proof}
Here we prove the theorem for the case where $\chi$ is a constant symmetric matrix, where condition (\ref{eq:2.8}) corresponds to $\chi$ being positive semi-definite. We refer the reader to \cite{Figotin:2006jy} for the proof of the general case. Consider the Hamiltonian system
\begin{subequations}
\begin{align}
		\label{eq:2.10.a} & \dot{z}(t) = \mathbb J_{2n} K^T f(t), \\
		\label{eq:2.10.b} & \partial_t \theta(t,x) = \partial_x^2 \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\chi}  f(t), \\
		\label{eq:2.10.c} & \partial_t \phi(t,x) = \theta(t,x),
\end{align}
\end{subequations}
together with the initial condition
\begin{equation} \label{eq:2.10.1}
	z(0) = z_0,\quad \theta(0,\cdot) = 0, \quad \phi(0,\cdot) = 0.
\end{equation}
Here $\theta$ and $\phi$ are vector valued functions in $\mathcal H^{2n}$, $\delta_0(s)$ is the Dirac's delta function, $\sqrt{ \chi}$ is the matrix square root of $\chi$ and $f$ is the solution to the equation
\begin{equation} \label{eq:2.11}
	f(t) + \sqrt{2} \cdot \sqrt{ \chi } \phi(t,0) = Kz(t).
\end{equation}
To show that the Hamiltonian system (\ref{eq:2.10.a}) to (\ref{eq:2.10.c}) is an extension to (\ref{eq:2.4}) in the sense discussed before, it is enough to show that the solution $f$ to the equation (\ref{eq:2.11}) also satisfies (\ref{eq:2.4.1}). Equations (\ref{eq:2.10.b}) and (\ref{eq:2.10.c}) are equations to a vibrating string, and can be solved analytically
\begin{equation} \label{eq:2.12}
	\phi(t,x) = \frac {\sqrt 2} 2 \int_0^{t-|x|} \sqrt{\chi} f(s)\ ds,\quad \theta(t,x) = \frac{\sqrt 2}{2} \cdot \sqrt{\chi} f(t - |x|).
\end{equation}
\end{proof}
We can recover (\ref{eq:2.4.1}) by substituting (\ref{eq:2.12}) into (\ref{eq:2.11}). The extended Hamiltonian $H_\text{ex}$ for the system (\ref{eq:2.10.a}) to (\ref{eq:2.10.c}) takes the quadratic from
\begin{equation} \label{eq:2.13}
	H_\text{ex}(z,\phi,\theta) = \frac 1 2 \left( \| Kz - \phi(t,0) \|_2^2 + \| \theta(t) \|^2_{\mathcal H^{2n} } + \| \partial_x\phi(t)\|^2_{\mathcal H^{2n} }\right)
\end{equation}
where $\| \cdot \|_2$ is the Euclidean norm on $\mathbb R^{2n}$ and $\| \cdot \|_{\mathcal H^{2n}}$ is the induced norm from the inner product on $\mathcal H^{2n}$.

@@

Note that time integration of (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) involves the integration of $f$ in (\ref{eq:2.12}). In general $f(t)$ must be stored and may cause serious storage limitation in long-time integration. However, we are interested solely in finding $z(t)$ which depends on $f(t)$, and $\phi(t,0)$ that is the integral of the history of $f(t)$. So by carefully choosing a quadrature rule that uses the same quadrature nodes as the time integrator we can avoid storing the history of $f$. For example for the trapezoidal rule, we recover the recursive relation
\begin{equation} \label{eq:2.14}
	\int_{0}^{t_n} f(s) \ ds \approx \frac{\Delta t}{2} f(t_n) + \frac{\Delta t}{2} f(t_{n-1}) + \int_{0}^{t_{n-1}} f(s) \ ds,
\end{equation}
where $\Delta t$ is the time discretization step. The recursive relation in (\ref{eq:2.12}) suggests that only storing the value of the integral term together with the state of $f$ in the previous time step is required to evaluate the integral for the new time step. For other interpolation based quadrature rules, we can construct similar recursive rules of the form
\begin{equation}
	\int_{0}^{t_n} f(s) \ ds \approx \sum_{i=0}^{k} \omega_i f(t_{n-i})  + \int_{0}^{t_{n-k}} f(s) \ ds
\end{equation}
for some quadrature weights $\omega_i$, $i=1,\dots,k$ with $k\ll n$. Thus, time integration of (\ref{eq:2.10.a})-(\ref{eq:2.10.c}), requires storing only $k$ evaluations of $f$.

### 2017-03-07 dissipative_hamiltonian/chapters/3.model.tex
Numerical result added

A heuristic approach to find a solution to (\ref{eq:3.4}) is proposed by \cite{Peng:2014di}, suggesting that the symplectic basis $A$ should take the form
\begin{equation} \label{eq:3.5}
	A = 
	\begin{pmatrix}
		\Phi & 0 \\
		0 & \Phi
	\end{pmatrix},
\end{equation}
where $\Phi \in \mathbb{R}^{n\times k}$ is an orthonormal matrix. By constructing the \emph{combined snapshot matrix}
\begin{equation} \label{eq:3.6}
	S_{\text{combined}} = [q_1,\dots,q_N,p_1,\dots,p_N], \quad (q_i^T,p_i^T)^T = z(t_i),
\end{equation}
we define $\Phi=[u_1,\dots,u_k]$, where $u_i$ is the $i$-th left singular vector of $S_{\text{combined}}$. The symplectic basis obtained this way is referred to as the \emph{Cotangent Lift}. It is shown in \cite{Peng:2014di} that among all symplectic bases of the form (\ref{eq:3.5}) the cotangent lift minimizes the projection error.

@@

We would like to point out that the transformation that transforms (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) to (\ref{eq:3.12.a})-(\ref{eq:3.12.c}) is given by
\begin{equation}
	\mathcal A = \begin{pmatrix}
		A& 0 \\
		0& A
	\end{pmatrix} : \mathbb R^{2n} \times \mathcal H^{2n} \to \mathbb R^{2k} \times \mathcal H^{2k}.
\end{equation}
This is a symplectic transformation, since $\mathcal A^T J_{2n} \mathcal A = J_{2k}$. Further, the hidden strings $(\tilde \phi,\tilde \theta)$

>>>>

A heuristic approach to find a solution to (\ref{eq:3.4}) is proposed by \cite{Peng:2014di}, suggesting that the symplectic basis $A$ should take the form
\begin{equation} \label{eq:3.5}
	A = 
	\begin{pmatrix}
		\Phi & 0 \\
		0 & \Phi
	\end{pmatrix},
\end{equation}
where $\Phi \in \mathbb{R}^{n\times k}$ is an orthonormal matrix. By constructing the \emph{combined snapshot matrix}
\begin{equation} \label{eq:3.6}
	S_{\text{combined}} = [q_1,\dots,q_N,p_1,\dots,p_N], \quad (q_i^T,p_i^T)^T = z(t_i),
\end{equation}
we define $\Phi=[u_1,\dots,u_k]$, where $u_i$ is the $i$-th left singular vector of $S_{\text{combined}}$. The symplectic basis obtained this way is referred to as the \emph{Cotangent Lift}. It is shown in \cite{Peng:2014di} that among all symplectic bases of the form (\ref{eq:3.5}) the cotangent lift minimizes the projection error. Note that the basis $A$ constructed using the cotangent lift is an ortho-symplectic basis.

@@

We point out that the transformation that transforms (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) to (\ref{eq:3.12.a})-(\ref{eq:3.12.c}) is given by
\begin{equation}
	\mathcal A = \begin{pmatrix}
		A& 0 \\
		0& A
	\end{pmatrix} : \mathbb R^{2n} \times \mathcal H^{2n} \to \mathbb R^{2k} \times \mathcal H^{2k}.
\end{equation}
This is a symplectic transformation, since $\mathcal A^T J_{2n} \mathcal A = J_{2k}$. Further, the dissipation of energy in the reduced system only depends on the reduced susceptibility. Thus, the choice of $A$ should be independent of the hidden strings $(\tilde \phi, \tilde \theta)$. In other words, if the reduced space is chosen to be a symplectic subspace, then the actions of model reduction and Hamiltonian extension commute. We summarize the algorithm for model reduction of dissipative Hamiltonian systems in Algorithm \ref{alg:3.1}.

\begin{algorithm}
\caption{Model Reduction For Dissipative Hamiltonian Systems} \label{alg:3.1}
\begin{enumerate}
	\item Construct the Hamiltonian extension (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) to the original TDD system (\ref{eq:2.4}).
	\item Collect the snapshots $z(t_i)$, $i=1,\dots,N$ through time integration of the extended Hamiltonian.
	\item Construct an ortho-symplectic basis $A$.
	\item Define $\tilde L = A^T L A$, $\tilde \chi = A^T \chi A$ and construct the reduced dissipative Hamiltonian system (\ref{eq:3.12.a})-(\ref{eq:3.12.c})
\end{enumerate}
\end{algorithm}

Note that Algorithm \ref{alg:3.1} does not depend on the choice of the method to construct an ortho-symplectic basis $A$. Thus, an SVD-based method, such as the cotangent lift, or a greedy-natured method can be applied.
